---
title: "Matrix algebra"
author: "Luis Fernando Delgado Munoz"
institution: "Alliance CIAT - Bioversity"
date: "2022-11-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Type of Matrices

A matrix is a rectangular array of elements. These elements are, as in a spreadsheet, arranged in different rows and columns. A matrix is denoted in uppercase and in bold typeface. The elements of a matrix are enclosed in brackets, e.g.,

\[
  A  =
  \left[ {\begin{array}{cc}
    {1} & {2} \\
    {3} & {4} \\
    {5} & {6} \\
  \end{array} } \right]
\]

This **A** matrix can be described as a 3 x 2 matrix, with the number of rows given first. The *ij*th element refers to that in the *i*th row and *j*th column of the matrix. Matrix elements are denoted in lower case followed by the row and column in subscript. In the example the matrix elements $a_{11} = 1$, $a_{12} = 2$, $a_{21} = 3$, $a_{22} = 4$, $a_{31} = 5$ and $a_{33} = 6$ 

A vector has only one row (i.e., **row vector**) or one column (i.e **column vector**). Vectors are denoted in lowercase and in **bold** typeface. An example of a vector is a **random vector**, in wich the elements of a random vector corresponds to the observed yields of three genotypes.

\[
\textbf{y} =
\left[{\begin{array}{cc}
{Y_1} \\
{Y_2} \\
{Y_3}
\end{array}}\right]
\]

A square matrix has equal number of rows and columns. There are special types of square matrices. In a **symetric matrix**, the element in the **i**th row and **j**th column is equal to the element in the **j**th row and **i**th column. An example of a **symetric** matrix is a **covariance matrix**, in wich the variances of the random variables are the **diagonal elements**, whereas the covariances between variables are the **off-diagonal element**. consider two variables, $Y_{1}$ and $Y_{2}$ with variances of $V(Y_{1}) = 10$ and $V(Y_{2}) = 20$ and the covariance of $Cov(Y_{1}, Y_{2}) = 5$. The covariance matrix of $Y_{1}$ and $Y_{2}$ is:

\[
  A  =
  \left[ {\begin{array}{cc}
    {V(Y_{1})} & {Cov(Y_{1}, Y_{2})} \\
    {Cov(Y_{1}, Y_{2})} & {V(Y_{2})} \\
  \end{array} } \right]
\]
\[
  =
  \left[ {\begin{array}{cc}
    {10} & {5} \\
    {5} & {20} \\
  \end{array} } \right]
\]

## Diagonal Matrix

If $Y_{1}$ and $Y_{2}$ are uncorrelated [i.e,. $Cov(y_{1}, Y_{2}) = 0$], the the covariance matrix becomes a **diagonal matrix** in wich all off-diagonal elements are equal to zero.

\[
 V (y) =
  \left[ {\begin{array}{cc}
    {10} & {0} \\
    {0} & {20} \\
  \end{array} } \right]
\]

## Identity Matrix

An **identity matrix**, which is denoted by **I** is a diagonal matrix with 1s, e.g,. 

\[
 I =
  \left[ {\begin{array}{cc}
    {1} & {0} \\
    {0} & {1} \\
  \end{array} } \right]
\]

## Null Matrix

A **Null matrix**, which is denoted by **O**, includes only 0s, e.g.,

\[
 O =
  \left[ {\begin{array}{cc}
    {0} & {0} \\
    {0} & {0} \\
  \end{array} } \right]
\]

# Matrix Operations

Next, I introduce some computational details of mixed models using matrix algebra.

## The Transposition of a matrix

The **Transposition of a matrix** is obtained by interchanging its rows and columns; the transpose of a matrix is denoted by the **′**. The result of a transpose matrix is another matrix with rows equal to the columns of the first matrix and columns equal to the rows of the first matrix, e.g.,

\[
  A  =
  \left[ {\begin{array}{cc}
    {1} & {2} \\
    {3} & {4} \\
    {5} & {6} \\
  \end{array} } \right]
\]
\[
  A′  =
  \left[ {\begin{array}{cc}
    {1} & {3} & {5} \\
    {2} & {4} & {6} \\
  \end{array} } \right]
\]

In other words, the element in the **j**th row and **i**th column of **A′** is $a′_{ji} = a_{ij}$.

### The R code of transposition of a matrix:

1. Load the needed libraries

```{r eval=FALSE}
install.packages("pacman") # Install the library
library(pacman) # load the library
pacman::p_load("Matrix", "reshape2") #It checks to see if a package is installed, if not it attempts to install the package
```

```{r}
pacman::p_load("Matrix", "reshape2") #It checks to see if a package is installed, if not it attempts to install the package
```

2. Create **A** matrix

```{r}
A = matrix(c(1, 2,
             3, 4,
             5, 6), nrow = 3, ncol = 2, byrow = T)
print(A)
```

3. Transpose the matrix **A**

```{r}
# t() function returns the transpose of A
A_transpose = t(A)
print(A_transpose)
```

## The product of a scalar and a matrix

The product of a scalar and a matrix is equal to the scalar multiplied by each element of the matrix. Suppose the variances of $Y_{1}$ and $Y_{2}$ are both equal to $V_{Y} = 10$, and the covariance between them is zero. The covariance matrix of $Y_{1}$ and $Y_{2}$ can the be expressed as the product of a scalar (i.e., $V_{Y}$) and an identity matrix:

$V(y) = \textbf{I}V_{Y}$
\[
      =
  \left[ {\begin{array}{cc}
    {1} & {0} \\
    {0} & {1} \\
  \end{array} } \right]10
\]
\[
      =
  \left[ {\begin{array}{cc}
    {10} & {0} \\
    {0} & {10} \\
  \end{array} } \right]
\]

### The R code of the product of a scalar and a matrix

1. Create the scalar
```{r}
V = 10
print(V)
```

2. Create the **identity matrix**

```{r}
# Use diag() function to create a 2 x 2 identity matrix
I <- diag(2)
print(I)
```

2. Product of a scalar and a matrix

```{r}
I*V
```

## The product of two matrices

the **product of two matrices**, $AB = C$, can be obtained only when the number of columns in $A$ is equal to the number of rows in $B$. Matrix $C$ the has the same number of rows as $A$ and the same number of columns as $B$:
$A_{mxn}B_{nxp} = C_{mxp}$

Where the dimensions of the matrices are denoted in parenthesis; the product of two matrices obtained by simple multiplying their correspondence elements. Instead, the element in the **i**th row and **j**th column of $C$ is obtained as the sum of the products of corresponding elements in the **i**th row of $A$ and **j**th column of $B$:

\[\sum_{s =1}^{10}a_{is}b_{sj}\]

To illustrate, suppose we wish to multiply $A$, which has $m =3$ rows and $n = 2$ columns, and $B$, which has $n = 2$ rows and $p = 2$ columns:

\[
    A =
  \left[ {\begin{array}{cc}
    {1} & {2} \\
    {3} & {4} \\
    {5} & {6} \\
  \end{array} } \right],
    B =
  \left[ {\begin{array}{cc}
    {1} & {0} \\
    {0} & {2} \\
   \end{array} } \right]
\]

We begin by obtaining the element in row 1 and column 1 of $C$ as:

\[c_{11} = a_{11}b_{11}+a_{12}b_{21}\]
\[= (1)(1) + (2)(0)\]
\[=1\]

This operation is illustrated by considering only the relevant elements in $A, B$ and $C$:

\[
   \left[ {\begin{array}{cc}
    {1} & {2} \\
    {.} & {.} \\
    {.} & {.} \\
  \end{array} } \right]
  \left[ {\begin{array}{cc}
    {1} & {.} \\
    {0} & {.} \\
   \end{array} } \right]\Rightarrow\left[ {\begin{array}{cc}
    {1} & {.} \\
    {.} & {.} \\
    {.} & {.} \\
  \end{array} } \right]
\]

The element in row 2 and column 1 of $C$ is obtained as

\[c_{21} = a_{21}b_{11}+a_{22}b_{21}\]
\[= (3)(1) + (4)(0)\]
\[=3\]

Again, this operation is illustrated by considering only the relevant elements in $A, B$ and $C$:

\[
  \left[ {\begin{array}{cc}
    {.} & {.} \\
    {3} & {4} \\
    {.} & {.} \\
  \end{array} } \right]
  \left[ {\begin{array}{cc}
    {1} & {.} \\
    {0} & {.} \\
   \end{array} } \right]\Rightarrow\left[ {\begin{array}{cc}
    {.} & {.} \\
    {3} & {.} \\
    {.} & {.} \\
  \end{array} } \right]
\]

This multiply-and-sum procedure is repeated until the last element is obtained as:

\[c_{32} = a_{31}b_{12}+a_{32}b_{22}\]
\[= (5)(0) + (6)(2)\]
\[=12\]

Matrix $C$ is then equal to:

\[
    C =
  \left[ {\begin{array}{cc}
    {1} & {4} \\
    {3} & {8} \\
    {5} & {12} \\
  \end{array} } \right]
\]

### The R code of the product of two matrices

1. Create the Matrix B

```{r}
B = matrix(c(1, 0,
             0, 2
             ), nrow = 2, ncol = 2, byrow = T)
print(B)
```

2. Product of Matrix $A$ and a matrix $B$

```{r}
# the operator %*% is used in the multiplication of matrices
C = A%*%B
print(C)
```

## The inverse of a square matrix

The inverse of a square matrix can by obtained as follows:
Suppose $A$ is a 2 x 2 matrix

\[
  A  =
  \left[ {\begin{array}{cc}
    {a_{11}} & {a_{12}} \\
    {a_{21}} & {a_{22}} \\
  \end{array} } \right]
\]

The inverse of this 2 x 2 matrix is obtained as:

\[A^{-1}=\frac{1}{a_{11}a_{22}-a_{12}a_{21}}\left[ {\begin{array}{cc}
    {a_{22}} & {-a_{12}} \\
    {-a_{21}} & {a_{11}} \\
  \end{array} } \right] 
\]

Lets explain more precisely with a numeric matrix:

Suppose $X$ is a 2 x 2 matrix:

\[
  x  =
  \left[ {\begin{array}{cc}
    {2} & {3} \\
    {3} & {4} \\
  \end{array} } \right]
\]

The inverse of this 2 x 2 matrix is obtained as:

\[x^{-1}=\frac{1}{(4)(2)-(3)(3)}\left[ {\begin{array}{cc}
    {4} & {-3} \\
    {-3} & {2} \\
  \end{array} } \right] 
\]
\[
  x^{-1}  = -1
  \left[ {\begin{array}{cc}
    {2} & {3} \\
    {3} & {4} \\
  \end{array} } \right]
\]
\[
  x^{-1}  =
  \left[ {\begin{array}{cc}
    {-4} & {3} \\
    {3} & {-2} \\
  \end{array} } \right]
\]


### The R code of inverse of a matrix

1. Create the Matrix x

```{r}
x = matrix(c(2, 3,
             3, 4
             ), nrow = 2, ncol = 2, byrow = T)
print(x)
```

2. Let's calculate the inverse of matrix x

```{r}
# Inverse of x is obtained using "solve" function
solve(x)
```

# bibliography

1. [Breeding for quantitative traits in plants](http://stemmapress.com/)
2. [Genetic data analysis for plant and animal breeding](https://link.springer.com/book/10.1007/978-3-319-55177-7)



